Model for DES Type Ia Supernova Cosmology Analysis

### 2015-07-24 doc/model.tex
more comments from Danny input

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
that have passed some selection criteria (e.g.\ classified as Type~Ia supernovae, SNe~Ia).  The probability distribution
function (pdf) of the Hubble diagram is denoted as
\begin{equation}
p({\mu},{z} |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S,\text{GC},
{\text{RA}}, {\text{Dec}}, \text{Detected}, {\text{Selected}}).
\label{hd:eqn}
\end{equation}
The set of distance moduli
and redshifts $(\mu, z)$ is the Hubble diagram.  The available information
from which the Hubble diagram is inferred include:
the transient photometry ${ADU}$; spectral measurements of
transient
type, redshift and SN-parameters ${T}_S$, ${z}_S$, ${\theta}_S$;
the galaxy catalog ${\text{GC}}$ and associated galaxy properties such as redshift,
mass, sSFR, etc.; transient coordinates  $\text{RA}$, $\text{Dec}$;
and the fact that the transients were Detected and Selected for the analysis.

The pdf in Equation~\ref{hd:eqn} is proportional to
\begin{equation}
p({\mu},{z}, \text{RA}, \text{Dec}, \text{GC},  \text{Detected}, {\text{Selected}}
 |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S)
\label{hd2:eqn}
\end{equation}
and this is what we model.  A sketch of the model (with the optional inclusion
of calibration) is depicted in the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.
Note that although we have been discussing the model for the Hubble diagram, the same PGM can
 be used to express the pdf for cosmological
parameters, labelled as $\theta_\mu$.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   The forward modeling
   flow goes from left to right, starting with the description of the universe, observatory,
   and data.
   A galaxy catalog GC informs the model for potential hosts $G$.  A transient
   at coordinates  $\text{RA/Dec}$  has a host $g_i$ drawn from $G$.
   The galaxy redshift redshift $z$ and parameters $\theta_{gi}$ depend on the host.
   A distance modulus model with parameters $\theta_\mu$ fix the distance modulus $\mu_i$.
   The transient type $T$ depends on the host-galaxy.   The transient
   parameters $\theta_T^X$, $\theta_{Ti}^X$ (and perhaps the host galaxy) fix the luminosity $L$.       The 
   incoming photon flux $n$, $n_g$  are then fixed
   with redshift and distance modulus.
   The instrumental transmission function $\phi$ is calibrated with data ${Z}$ and
   gives the expected
   counts $\overline{\mathit{ADU}}$, $\overline{\mathit{ADU}}_g$. 
   The transient must be Detected and Selected and its realized light curve (${ADU}$) depends on the expected counts.  Some transients have spectral data
   ${X}_S$.
   \label{pgm:fig}}
\end{figure}

The model describes the transients that constitute the cosmology analysis, given the following
information.  
\begin{itemize}
\item ${\text{RA}}$, ${\text{Dec}}$.  The observed
coordinates.  For our purposes the observed, rather than true, coordinates are sufficient.
\item Detected.  The transient is discovered.
\item Selected. The transient is selected for the cosmology analysis.  This presumably
includes its classification as SN~Ia.
\item GC.  The galaxy catalog used to make host-galaxy associations, including with
redshifts and galaxy properties.
\end{itemize}

The universe is modeled as the background cosmology, transients that could
enter the sample, and their host galaxies.
\begin{itemize}
\item $G$ is a shorthand for the model of all galaxies that can host a transient.  It is
the union of the objects in the galaxy catalog GC
plus unobserved galaxies that have not
met the detection threshold, presumably due to their low surface brightness.
\item $P(g_i | G, {\text{RA}}/{\text{Dec}})$. The host galaxy $g_i$
is drawn from the GC based on coordinates.
This pdf is
the deliverable we want from Ravi.
\item $z_i$, $\theta_{gi}$ are the galaxy redshift and parameters (e.g.\ mass, sSFR, metallicity).
For observed galaxies the parameter values are based on measurements e.g.\ $\theta_{gi} \sim \mathcal{N}({\theta}_{gi},\sigma^2_{{\theta}_{gi}})$, otherwise the distribution
of $\theta_{gi}$ for ``hostless'' galaxies is described by the model.
Redshifts may be photometric or spectroscopic.
\item $\theta_\mu$ are the parameters that specify that distance modulus.
\item $\mu_i=\mu(z_i; \theta_\mu)$.  The distance modulus is a function of redshift,
whose specific form is unspecified at the moment,  it could
be a $\Lambda$CDM prediction with cosmological parameters, a spline with knot values
as parameters.
The redshift of the galaxy is used in the Hubble diagram.
\item $P(T_i | \theta_{gi})$.  The transient type $T_i$, SN~Ia or non-Ia, depends
on the host properties.  We could make the relative rates be dependent
on specific star formation rate and morphology.
\item $p(\theta_T^{Ia}, \theta_{Ti}^{Ia})$.  The model parameters for SNe~Ia, including
parameters that describe distributions.
Different SN~Ia models could be considered: the treatment of SALT2 parameters
has been presented in \citet{2011MNRAS.418.2308M}.
The SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would constitute $\theta_{Ti}^{Ia}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would constitute $\theta_T^{Ia}$.
\item $p(\theta_T^{non-Ia}, \theta_{Ti}^{non-Ia})$.  Non-Ia transients
that can be mistyped as Ia are described by a parameterized model.
\item $L_i(t,\lambda; T_i, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia},
\theta_{gi})$.  The source model fixes
the luminosity. The  model includes  information on the
source and line-of-sight effects, which specify the effective SED.   Only the
parameters of the model of the appropriate type $T$ are relevant.  The galaxy parameters
$\theta_{gi}$ are included if correlations with host-galaxy properties are to be modeled.
\item $n_i(t,\lambda; L_i, \mu_i, z_i)$, $n_{gi}(t,\lambda; L_i, \mu_i, z_i)$.  The  fluxes of
the transient and host that are incident at Earth
are fixed by the luminosity, distance modulus, and redshift.
\end{itemize}

@@

The observations are as follows:
\begin{itemize}
\item ${\mathit{ADU}}$ is the realized flux can be expressed as
\begin{equation}
P({\mathit{ADU}} | \text{Dectected}, \text{Selected}, \overline{\mathit{ADU}}, 
\overline{\mathit{ADU}}_g)=\frac{P({\mathit{ADU}},\text{Dectected}, \text{Selected} |  
\overline{\mathit{ADU}},  \overline{\mathit{ADU}}_g)}{P(\text{Dectected}, \text{Selected} | 
\overline{\mathit{ADU}},  \overline{{\mathit{ADU}}}_g)},
\end{equation}
which can be determined from detection and typing efficiency studies.
\item ${T}_S$, ${z}_S$, ${\theta}_S$ Transient type, redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
A very simple association is ${T}_S = T$, ${z}_S=z$, ${\theta}_S \in \theta_g$,
or a blurry version instead of a straight equality.
\item ${Z}$ represents the calibration measurements.  Although we expect to receive
the transmission functions from the Calibration Task Force, we may be want to include
calibration as part of our analysis.
\end{itemize}

>>>>

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
that have passed some selection criteria (e.g.\ classified as Type~Ia supernovae, SNe~Ia).  The probability distribution
function (pdf) of the Hubble diagram is denoted as
\begin{equation}
p({\mu},{z} |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S; \text{GC},
{\text{RA}}, {\text{Dec}}, \text{Detected}, {\text{Selected}}).
\label{hd:eqn}
\end{equation}
The set of distance moduli
and redshifts $(\mu, z)$ is the Hubble diagram.  The available information
from which the Hubble diagram is inferred include:
the transient photometry ${ADU}$; spectral measurements of
transient
type, redshift and SN-parameters ${T}_S$, ${z}_S$, ${\theta}_S$;
the galaxy catalog ${\text{GC}}$ and associated galaxy properties such as redshift,
mass, sSFR, etc.; transient coordinates  $\text{RA}$, $\text{Dec}$;
and the fact that the transients were Detected and Selected for the analysis.

The pdf in Equation~\ref{hd:eqn} is proportional to
\begin{equation}
p({\mu},{z}; \text{RA}, \text{Dec}, \text{GC},  \text{Detected}, {\text{Selected}}
 |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S)
\label{hd2:eqn}
\end{equation}
and this is what we model.  A sketch of the model (with the optional inclusion
of calibration) is depicted in the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.
Note that although we have been discussing the model for the Hubble diagram, the same PGM can
 be used to express the pdf for cosmological
parameters, labelled as $\theta_\mu$.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   The forward modeling
   flow goes from left to right, starting with the description of the universe, observatory,
   and data.  The galaxies that host transients are described by a galaxy catalog GC. 
   A transient has
   coordinates  $\text{RA/Dec}$.  Its host galaxy $g_i$ is drawn from from the GC.
   The galaxy redshift redshift $z$ and parameters $\theta_{gi}$ depend on the host.
   A distance modulus model with parameters $\theta_\mu$ fix the distance modulus $\mu_i$.
   The transient type $T$ depends on the host-galaxy.   The transient
   parameters $\theta_T^X$, $\theta_{Ti}^X$ (and perhaps the host galaxy) fix the luminosity $L$.       The 
   incoming photon flux $n$, $n_g$  are then fixed
   with redshift and distance modulus.
   The instrumental transmission function $\phi$ is calibrated with data ${Z}$ and
   gives the expected
   counts $\overline{\mathit{ADU}}$, $\overline{\mathit{ADU}}_g$. 
   The transient must be Detected and Selected and its realized light curve (${ADU}$) depends on the expected counts.  Some transients have spectral data
   ${X}_S$.
   \label{pgm:fig}}
\end{figure}

The model describes the transients that constitute the cosmology analysis, given the following
information.  
\begin{itemize}
\item ${\text{RA}}$, ${\text{Dec}}$.  The observed
coordinates.  For our purposes the observed, rather than true, coordinates are sufficient.
\item Detected.  The transient is discovered.
\item Selected. The transient is selected for the cosmology analysis.  This presumably
includes its classification as SN~Ia.
\item GC.  The galaxy catalog used to make host-galaxy associations, including with
redshifts and galaxy properties.  It is
the union of the objects in the galaxy catalog GC
plus unobserved galaxies that have not
met the detection threshold, presumably due to their low surface brightness.
\end{itemize}

The universe is modeled as the background cosmology, transients that could
enter the sample, and their host galaxies.
\begin{itemize}
\item $P(g_i ; \text{GC}, {\text{RA}}/{\text{Dec}})$. The host galaxy $g_i$
is drawn from the GC based on coordinates.
This pdf is
the deliverable we want from Ravi.
\item $z_i$, $\theta_{gi}$ are the galaxy redshift and parameters (e.g.\ mass, sSFR, metallicity).
For observed galaxies the parameter values are based on measurements e.g.\ $\theta_{gi} \sim \mathcal{N}({\theta}_{gi},\sigma^2_{{\theta}_{gi}})$, otherwise the distribution
of $\theta_{gi}$ for ``hostless'' galaxies is described by the model.
Redshifts may be photometric or spectroscopic.
\item $\theta_\mu$ are the parameters that specify that distance modulus.
\item $\mu_i=\mu(z_i; \theta_\mu)$.  The distance modulus is a function of redshift,
whose specific form is unspecified at the moment,  it could
be a $\Lambda$CDM prediction with cosmological parameters, a spline with knot values
as parameters.
The redshift of the galaxy is used in the Hubble diagram.
\item $P(T_i | \theta_{gi})$.  The transient type $T_i$, SN~Ia or non-Ia, depends
on the host properties.  We could make the relative rates be dependent
on specific star formation rate and morphology.
\item $p(\theta_T^{Ia}, \theta_{Ti}^{Ia})$.  The model parameters for SNe~Ia, including
parameters that describe distributions.
Different SN~Ia models could be considered: the treatment of SALT2 parameters
has been presented in \citet{2011MNRAS.418.2308M}.
The SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would constitute $\theta_{Ti}^{Ia}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would constitute $\theta_T^{Ia}$.
\item $p(\theta_T^{non-Ia}, \theta_{Ti}^{non-Ia})$.  Non-Ia transients
that can be mistyped as Ia are described by a parameterized model.
\item $L_i(t,\lambda; T_i, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia},
\theta_{gi})$.  The source model fixes
the luminosity. The  model includes  information on the
source and line-of-sight effects, which specify the effective SED.   Only the
parameters of the model of the appropriate type $T$ are relevant.  The galaxy parameters
$\theta_{gi}$ are included if correlations with host-galaxy properties are to be modeled.
\item $n_i(t,\lambda; L_i, \mu_i, z_i)$, $n_{gi}(t,\lambda; L_i, \mu_i, z_i)$.  The  fluxes of
the transient and host that are incident at Earth
are fixed by the luminosity, distance modulus, and redshift.
\end{itemize}

@@

The observations are as follows:
\begin{itemize}
\item ${\mathit{ADU}}$ is the realized flux can be expressed as
\begin{equation}
P({\mathit{ADU}} | \overline{\mathit{ADU}}, 
\overline{\mathit{ADU}}_g; \text{Dectected}, \text{Selected})=\frac{P({\mathit{ADU}};\text{Dectected}, \text{Selected} |  
\overline{\mathit{ADU}}, \overline{\mathit{ADU}}_g)}{P(\text{Dectected}, \text{Selected} | 
\overline{\mathit{ADU}},  \overline{{\mathit{ADU}}}_g)},
\end{equation}
which can be determined from detection and typing efficiency studies.
\item ${T}_S$, ${z}_S$, ${\theta}_S$ Transient type, redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
A very simple association is ${T}_S = T$, ${z}_S=z$, ${\theta}_S \in \theta_g$,
or a blurry version instead of a straight equality.
\item ${Z}$ represents the calibration measurements.  Although we expect to receive
the transmission functions from the Calibration Task Force, we may be want to include
calibration as part of our analysis.
\end{itemize}
